Dynamic capabilities and competitive advantage

### 2019-11-01 article4_jbr.rnw
added buyin and incentinve alignment to theory

DC as resource orchestration - influencing capabilities
DC as cognitive programmer - inluencing cognitive frames

@@

Looking at dynamic capabilities through the functions of sensing and seizing opportunities
\cite{Teece2007} makes an obvious case for resource orchestration which is also
highlighted in the extant literature. This function is related to the discovering and
capturing opportunities (e.g. the ability of the organization to sense and seize
opportunities (ibid)) which is very much related to the concept of exploration
\citep{March1991}Moreover, as discussed above, these functions are also requiring a
certain non-behavioral object of action in order to be fully efficient. To achieve a
competitive edge firms will have to be able to sense and seize, which require a certain
organizational cognitive frame to be efficient. 'Dynamic capabilities are about doing the
right things, at the right time, based on new product (and process) development, unique
managerial orchestration processes, a strong and {\bf change-oriented organizational
  culture}' (our emphasis) \cite[p]{Teece2014}.

changes resulting from execution of dynamic capabilities (e.g. sensing and seizing) may
move existing organizational objective functions and thus render current incentive
alignment obsolete or out of place. This leads to an objective displacement that is
particularly prevalent in firms with high sensing and seizing capabilities because distant
search may lead to more radical change (ref). Organizations with high levels of dynamic
capabilities geared towards sensing and seizing opportunities are doing so through
\emph{resource orchestration}, e.g. by creating, extending and modifying the resource
base, as discussed extensively in the extant literature
\citep{Helfat2007,DiStefano2014,Protogerou2012,Schilke2018}. However, sensing and seizing
opportunities in the market relates to discovering market disequilibrium
\citep{Kirzner1997} as well as upsetting existing equilibria \citep{Schumpeter1934}. The
process of sensing and seizing 'involves understanding latent demand, the structural
evolution of industries and markets, and likely supplier and competitor responses'
\citep[p. 1322]{Teece2007} as well as 'a good deal of customer, competitor, and supplier
information and intelligence (with a) a significant tacit component' (ibid p 1330).To make
sure the organization respond to changes (i.e. having an efficient cognitive frame)
stemming from executed dynamic capabilities (e.g. sensing and seizing) the firm needs to
make sure incentives are aligned for change.

Through the process of exploration, individuals in the organization may
accumulate private and tacit information that, if unchecked, can become a source of
knowledge asymmetries and incentive problems (alchian and demsetz 1972), and leave the
firm vulnerable to agency problems. Accumulation of private information in the
organization without incentive alignment of some sort can lead rent-appropriation from
information asymmetries \citep{Blyler2003}. In such instances of potential misalignment of
interest, a shared cognitive frame emphasizing a growth mindset can be a useful
remedy. Furthermore, in addition to incentives alignment the more psychological element
of change buy-in comes into play \todo{Not sure how to proceed here - should we make a
  distinction between the incentive problem and the buy in problem? Coud they both be
  examples of interest alignment with different faces of motivation - extrensic vs intrinsic} (ref).

To see how this would work recent contribution in neuroscience is helpful. Several
contributions have established a clear connection between having a growth mindset and
emphasizing intrinsic motivation (Schroder et.al 2014,O`Rourke et.al 2014). Moreover,
individuals with higher degrees of intrinsic motivation from a growth mindset exhibit a
significant higher ability for learning and consequently for embracing new learning
opportunities (Yeager et al 2012). In other words, individuals with a growth mindset are
more open to change because it enhances their intrinsic motivation from learning which
fuels their individual utility (Ng 2018). In a situation with considerable change stemming
from execution of dynamic capabilities, an organization with a common cognitive frame
based on a growth mindset will be able to capture more value from the change \emph{et
  ceteris paribus}.

@@

>>>>

DC as resource orchestration - influencing capabilitie
DC as cognitive programmer - inluencing cognitive frames

@@

Looking at dynamic capabilities through the functions of sensing and seizing opportunities
\cite{Teece2007} makes an obvious case for resource orchestration which is also
highlighted in the extant literature. This function is related to the discovering and
capturing opportunities (e.g. the ability of the organization to sense and seize
opportunities (ibid)) which is very much related to the concept of exploration
\citep{March1991}. Moreover, as discussed above, these functions are also requiring a
certain non-behavioral object of action in order to be fully efficient. To achieve a
competitive edge firms will have to be able to sense and seize, which require a certain
organizational cognitive frame to be efficient. 'Dynamic capabilities are about doing the
right things, at the right time, based on new product (and process) development, unique
managerial orchestration processes, a strong and {\bf change-oriented organizational
  culture}' (our emphasis) \cite[p]{Teece2014}.

changes resulting from execution of dynamic capabilities (e.g. sensing and seizing) may
move existing organizational objective functions and thus render current incentive
alignment obsolete or out of place. This leads to an objective displacement that is
particularly prevalent in firms with high sensing and seizing capabilities because distant
search may lead to more radical change (ref). Organizations with high levels of dynamic
capabilities geared towards sensing and seizing opportunities are doing so through
\emph{resource orchestration}, e.g. by creating, extending and modifying the resource
base, as discussed extensively in the extant literature
\citep{Helfat2007,DiStefano2014,Protogerou2012,Schilke2018}. However, sensing and seizing
opportunities in the market relates to discovering market disequilibrium
\citep{Kirzner1997} as well as upsetting existing equilibria \citep{Schumpeter1934}. The
process of sensing and seizing 'involves understanding latent demand, the structural
evolution of industries and markets, and likely supplier and competitor responses'
\citep[p. 1322]{Teece2007} as well as 'a good deal of customer, competitor, and supplier
information and intelligence (with a) a significant tacit component' (ibid p 1330).To make
sure the organization respond to changes (i.e. having an efficient cognitive frame)
stemming from executed dynamic capabilities (e.g. sensing and seizing) the firm needs to
make sure incentives are aligned for change and that the organization buys into the changes.

First, the \emph{incentive alignment problem} relates to how interests of the firm
as a whole and the collection of individual are aligned \cite{Gottschalg2007} and how this
alignment changes when the firm embarks on strategic change as triggered by execution of
dynamic capabilities. Through the process of exploration, individuals in the organization
may accumulate private and tacit information that, if unchecked, can become a source of
knowledge asymmetries and incentive problems (alchian and demsetz 1972), and leave the
firm vulnerable to agency problems. Accumulation of private information in the
organization without incentive alignment of some sort can lead rent-appropriation from
information asymmetries \citep{Blyler2003}. In such instances of potential misalignment of
interest, a shared cognitive frame emphasizing a growth mindset can be a useful remedy.

Second, the \emph{buy-in problem} refers to the general willingness of the organization to
accept and value the importance of some decision or system 'without being prompted or
required by formal governance mechanisms'\citep[p. 812]{Simsek2009}. When individuals of
a firm buys into changes they will enhance the full organization`s ability to not only act
upon changes (Marchand et.al 2000), but also in 'timely responsiveness' (Kohli et.al 1993,
Simsek2009).

Both the incentive alignment problem and the buy-in problem are interrelated, but still
represents different mechanisms. They are tied to individual motivation albeit in
different ways. The incentive alignment problem is tied to \emph{extrinsic motivation}
(Lindenberg 2001, Brief and Aldag 1977) where the individual is motivated by rewards or
outcome 'such as money, power, recognition and so forth'
\citep[p. 420]{Gottschalg2007}. The buy-in problem, in turn, is related to \emph{intrinsic
  motivation} where the individual is driven by the goal of being engaged in enjoyable
(Lindenberg, 2001), self-determined, and competence-enhancing (Deci \& Ryan, 1985)
behavior (hedonistic intrinsic motivation), as well as adhering to a goal of engaging in
behavior that is compli- ant with norms and values (normative intrinsic
motivation). \todo{some direct citations here - need rewrite}.

Extrinsic motivation is often addressed through reward systems (ref) but tacit mechanisms
such as psychological ownership \citep{Sieger2013} has been found to be very efficient in
aligning incentives. Psychological ownership basically retains agency theory’s assumption
of the self-interested manager, whereas the non-economic functions of ownership, such as
efficacy, identity, and territoriality, might curtail expropriating behaviour and align
the interests of agents and principals (cf. Pierce et al., 2003). Hence, this concept of
psychological ownership bridges over to the intrinsic motivation of individuals. Moreover,
intrinsic motivation in general is influenced through changing the tasks of individuals or
at the very least accepting individuals to change ttasks themselves to fit their
perception of their context (Hackman and Gersick), and through regimes of socialization
and shared norms (Kerr and Jackofsky). Common for both psychological ownership and the
intrinsic motivation enhancement in general is a feature of the employees' cognitive
framing, which guides the mental pathways by which they decide what is, and what is not,
important. According to Leonard- Barton (1992), these norms and values represent the
accretion of decisions made over time in response to the interpretation of organizational
roles. Consequently, a cognitive frame that enhances the intrinsic motivation of the
employees are very valuable. One way of thinking of instilling an appropriate and
efficient cognitive frame would be the concept of a growth mindset (Dweck) as also
emphasized in the case of Microsoft.

To see how this would work recent contribution in neuroscience is helpful. Several
contributions have established a clear connection between having a growth mindset (Dweck 200x) and
emphasizing intrinsic motivation (Schroder et.al 2014,O`Rourke et.al 2014). Moreover,
individuals with higher degrees of intrinsic motivation from a growth mindset exhibit a
significant higher ability for learning and consequently for embracing new learning
opportunities (Yeager et al 2012). In other words, individuals with a growth mindset are
more open to change because it enhances their intrinsic motivation from learning which
fuels their individual utility (Ng 2018). In a situation with considerable change stemming
from execution of dynamic capabilities, an organization with a common cognitive frame
based on a growth mindset will be able to capture more value from the change \emph{et
  ceteris paribus}.

@@

the general capability, views, norms and rules of behavior, with regard to accessing,
understanding and using information in a social collectivity. Zheng 2005
